Sequential algorithms to split and merge ultra-high resolution 3D images

### 2017-08-01 paper.tex
A better intro, still not perfect

\begin{abstract}
  Splitting and merging data is a requirement for most parallel
    or distributed processing operations, however, naive algorithms
    perform very poorly on ultra-high resolution 3D images due to
    important seek times. To address this issue, we created two
    algorithms based on in-memory reconstruction of image segments and
    analyzed their complexity in terms of number of seeks. Evaluation
    on a 3000x3000x3000 image of the brain shows that both our
    algorithms can split or merge blocks with a performance similar to
    slices, which solves our initial problem. We conclude that
    splitting and merging large 3D images can be done efficiently
    based on simple data formats.
\end{abstract}

@@

Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. BigBrain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. Other
examples found in medical imaging, our primary domain of interest,
include high-resolution 3D electron microscopy (see, e.g.,
~\cite{bock2011network}) or micro and
nano-tomography~\cite{10.1371/journal.pone.0035691}. As such images
would typically be processed on a computing cluster, possibly using
locality-aware file systems such as HDFS, software libraries are
needed to split and merge such images efficiently. We introduce and
compare a family of algorithms to address this problem in a sequential
environment made of a single storage disk. We are aware of the need
for parallel algorithms that would split and merge images to and from
an array of disks but we plan to describe such algorithms in a
subsequent paper based on the results of the present one. \todo{ There
  is BigBrain-1 rescanned at 1 micron in the plane. So resolution is
  1x1x20, 7404 slices. If raw data are about 1Tb, then multiply by 100
  for an estimation of the size of the new image. }

%also in
%MB/s).

Images are split into chunks representing 3D blocks or 2D slices. We
assume that a block or slice fits in memory. The BigBrain would
perhaps be split into 125 chunks of 600~MB. The decision to split an
image into slices or blocks, and the size of the resulting chunks, may
be constrained by the application. Some applications, for instance
spatial filtering, would commonly require blocks while other ones such
as acquisition artifact removal would rather work on
slices. Flexibility is thus required in the splitting
scheme. Applications processing voxels individually, for instance
histogram computation or k-means clustering, could work on either
slices or blocks.

The main problem encountered in the context of this paper is the fact
that, if not ordered properly, data reads and writes might result in
extensive seek times that, as we illustrate below, drastically limit
the I/O performance. This problem is obviously related to the ordering
of data bytes in image files, which we assume is arbitrary but known
to the algorithm. The problem is also related to the geometry of the
chunks to extract or merge, which we again assume arbitrary but known
to the algorithm. In a nutshell, we are looking for algorithms to
reduce seek times while allowing for arbitrary chunk geometries and
arbitrary byte orders in data files. The main idea of the variations
presented in the remainder of the paper is to convert data orderings
in memory before writing them on disk.

The literature is remarkably scarce on the problem described
above. Parallel processing of 3D images has obviously been extensively
studied, but methods have focused more on geometrical approaches to
partition images and on load-balancing and task scheduling techniques
while we aim at algorithms to efficiently split or merge images
regardless of the geometry of the chunks \todo{do a serious literature
review on this topic}. Perhaps this lack of related works could be
attributed to the fact that disk I/O weren't a bottleneck until image
resolution reached a certain threshold. A relevant reference is the
description of the Open Connectome Data Cluster~\cite{burns2013open},
a data warehouse system that allows users to retrieve specific chunks
of large datasets. In this work, the problem of long seek times is
addressed by designing a specific file format based on space-filling
curves, that elegantly preserves spatial proximity on disk. However,
addressing the problem using a specific data organization optimizes
the system for specific chunk geometries, volumetric in this case,
while we aim at algorithms that support multiple types of geometries.

This paper makes the following contributions:
\begin{itemize}
  \item We propose a set of algorithms to split and merge 3D images
    that do not fit in memory.
  \item We model the performance (number of seeks) of these algorithms,
    based on a simple but realistic disk I/O model.
  \item We evaluate our model and algorithms on a representative
    use-case involving a 1~TB image a hard drive and a solid-state
    drive.
\end{itemize}
Section~\ref{sec:algos} presents our algorithms and their complexity,
Section~\ref{sec:results} details the results and
Sections~\ref{sec:discussion} discusses them.

@@

\todo{Overlapping blocks. non-cubic blocks.}

@@

Without loss of generality, we consider a file format where voxels are
written after alphanumerical sorting of their (x,y,z)
representation. That is, x slices are written one after the other, and
in every x slice, y rows are written one after the other. NIfTI, the
file format defined by the Neuroimaging Informatics Technology
Initiative NIfTI\footnote{\url{https://nifti.nimh.nih.gov}} is such a
format, used by the vast majority of the neuroimaging community. In
the remainder, we use the term ``slice" for ``x slice".

@@

For the sake of mathematical simplicity, we assume that
the reconstructed image and blocks are cubic. Under this assumption,
we also have the following relations:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fill a row of the reconstructed
  image: $\sqrt[3]{n}$.
\end{itemize}
%\todo{Make a figure to illustrate the notations.}

@@

We use a simplistic disk model where a disk is characterized by its
read and write rates and its access time. In particular, we assume
that seeks require a constant amount of time, regardless of the
position seeked to. In practice, large variations would be expected,
but modeling such variations would inevitably lead to models specific
to the hardware, file system or operating system, which we
intentionally avoid here. Thus, our goal is to find algorithms that
minimize the \emph{number} of seek operations. Likewise, in modern
systems, read and write times are greatly impacted by caching
occurring at several levels, which we do not model here.

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks} are the
naive merging methods for slices and blocks.
\begin{algorithm}[h]
\caption{Naive merging from slices.}
\label{algo:naive-slices} 
\begin{algorithmic}
  \FOR{each slice}
    \STATE read slice
    \STATE write slice in reconstructed image
  \ENDFOR      
\end{algorithmic}
\end{algorithm}

\begin{algorithm}[h]
\caption{Naive merging from blocks.}
\label{algo:naive-blocks}
\begin{algorithmic}
  \FOR{each block}
    \STATE read block
    \STATE write block in reconstructed image
  \ENDFOR 
\end{algorithmic}
\end{algorithm}

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks}
actually have very different performance even though blocks and slices
have identical sizes. Since slices are stored contiguously in the
reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$:
\begin{equation}
  N_\mathrm{slices} = 2n 
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block. That is, the total number of seeks
performed by Algorithm~\ref{algo:naive-blocks} is:
\begin{equation}
N_\mathrm{blocks} = n+nd^2 
\end{equation}
In
practice, this difference could lead to a tremendous slowdown as we will show
later.

@@

The algorithms presented up to now are in fact particular cases of
memory buffering where the amount of available memory equals the
maximum size of a chunk.

\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE list = sort slices by increasing x values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = list[index]
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}

\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}}
\end{equation}

@@

\subsection{Data}
To benchmark our library we use the 76~GB BigBrain image split in 125
non-overlapping chunks. The data is available \todo{[here]}.

>>>>

\begin{abstract}
  Splitting and merging data is a requirement for most parallel or
  distributed processing operations. Naive algorithms, however,
  perform very poorly to split or merge 3D blocks from ultra-high
  resolution images, due to important seek times. We propose and
  analyze two algorithms based on in-memory reconstruction of image
  parts. Cluster reads access image splits only once while Multiple
  reads minimize the accesses to the image. Evaluation on a
  3000x3000x3000 brain image shows that our algorithms perform
  similarly to the optimal configuration (slice merging) provided that
  enough memory is available, which solves our initial problem.  We
  conclude that splitting and merging large 3D images can be done
  efficiently based on simple data formats.
\end{abstract}

@@

% Problem definition
Three-dimensional images that exceed typical memory size are
increasingly found in a variety of disciplines. Big~Brain, for
instance, is a 3D histological image of the human brain that
represents 1~TB of data organized in 3600 planes at full resolution
(1$\mu$m in-plane), and 76~GB at a 40$\mu$m isotropic resolution
commonly used in neurosciences~\cite{amunts2013bigbrain}. This dataset
is being re-scanned at an even higher resolution, leading to 100~TB of
raw data. Other examples found in medical imaging, our primary domain
of interest, include high-resolution 3D electron microscopy (see,
e.g., ~\cite{bock2011network}) or micro- and
nano-tomography~\cite{10.1371/journal.pone.0035691}. As such images
would typically be processed on a computing cluster, possibly using
locality-aware file systems such as HDFS~\cite{shvachko2010hadoop},
efficient software libraries are needed to split and merge them
efficiently. In particular, data reads and writes might result in
extensive seek times that drastically limit I/O performance.  In this
paper we introduce and compare a family of sequential algorithms to
split and merge images with minimal seek time.

Images are split into chunks representing 3D blocks or 2D slices,
which we assume fit in memory. A dataset such as Big~Brain would
perhaps be split into 125 chunks of 600~MB. The decision to split an
image into slices or blocks, and the size of the resulting chunks, is
usually done by the application. Some applications would commonly
require blocks, for instance spatial filtering, while other ones such
as acquisition artifact removal would rather work on
slices. Applications processing voxels individually, for instance
histogram computation or k-means clustering, could work on either
slices or blocks. Flexibility is thus required in the splitting
scheme.

Seek time is highly dependent on the byte ordering in image files,
which we assume is arbitrary but known to the algorithm. The geometry
of the chunks to split or merge (slices or blocks) and their sizes are
also critical, which we again assume arbitrary but known to the
algorithm.  The main idea of our algorithms will be to convert data
orderings in memory before writing them to disk.

The literature on this problem is remarkably scarce. Parallel and
distributed image processing has obviously been extensively studied
and used in various
platforms~\cite{miguet1991elastic,tang2007eman2,yang2008parallel,braunl2013parallel,moise2013terabyte,bajcsy2013terabyte},
but methods have focused more on geometrical approaches to partition
images and on load-balancing and task scheduling techniques while we
aim at algorithms to efficiently split or merge images regardless of
the geometry of the chunks. Seek times are, however, often identified
as an issue but the preferred solution is usually to optimize data
storage formats for a certain application while we are searching for
algorithms that would reduce the seek time regardless of the data
format. For instance, the Open Connectome Data
Cluster~\cite{burns2013open} is a data warehouse system that allows
users to retrieve specific chunks of large image datasets. It
addresses seek times by designing a specific file format based on
space-filling curves, which elegantly preserves spatial proximity on
disk. Such specific data formats optimize the system for particular
chunk geometries, blocks in this case, while we aim at algorithms to
support multiple types of geometries.
%http://dl.acm.org/citation.cfm?id=218395 also minimizes seek time using a specific format.

This paper makes the following contributions:
\begin{itemize}
  \item We propose a set of algorithms to split and merge
    terabyte-scale 3D images from blocks or slices.
  \item We model the performance (number of seeks) of these algorithms.
  \item We evaluate our model and algorithms on a representative
    use-case involving a 1~TB image a hard drive and a solid-state
    drive.
\end{itemize}
Section~\ref{sec:algos} presents our algorithms and their complexity,
Section~\ref{sec:results} details the results and
Sections~\ref{sec:discussion} discusses them.\todo{Revise when outline is final.}

@@

@@

Without loss of generality, we consider a file format where voxels are
written after alphanumerical sorting of their (z,y,x)
representation. That is, z slices are written one after the other, and
in every z slice, y rows are written one after the other. x is the
fastest-changing dimention.  NIfTI, the file format defined by the
Neuroimaging Informatics Technology Initiative
NIfTI\footnote{\url{https://nifti.nimh.nih.gov}} is such a format,
used by the vast majority of the neuroimaging community. In the
remainder, we use the term ``slice" for ``z slice".

@@

For the sake of mathematical simplicity, we assume that the
reconstructed image and blocks are cubic and that they do not
overlap. Under this assumption, we also have the following relations:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fill a row of the reconstructed
  image: $\sqrt[3]{n}$.
\end{itemize}

%\todo{Make a figure to illustrate the notations.}

@@

We use a simplistic disk model where a disk is characterized by its
read and write rates and its seek time. We do not distinguish between
access time and seek time. We also assume that seeks require a
constant amount of time, regardless of the position seeked to. In
practice, large variations would be expected, but modeling such
variations would inevitably lead to models specific to the hardware,
file system or operating system, which we intentionally avoid
here. Thus, our goal is to find algorithms that minimize the
\emph{number} of seek operations. Likewise, in modern systems, read
and write times are greatly impacted by caching occurring at several
levels, which we do not model here.

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks} are the
naive merging methods for slices and blocks.
\begin{algorithm}[h]
\caption{Naive merging from slices.}
\label{algo:naive-slices} 
\begin{algorithmic}
  \FOR{each slice}
    \STATE read slice
    \STATE write slice in reconstructed image
  \ENDFOR      
\end{algorithmic}
\end{algorithm}
\begin{algorithm}[h]
\caption{Naive merging from blocks.}
\label{algo:naive-blocks}
\begin{algorithmic}
  \FOR{each block}
    \STATE read block
    \STATE write block in reconstructed image
  \ENDFOR 
\end{algorithmic}
\end{algorithm}
Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks}
actually have very different performance even though blocks and slices
have identical sizes. Since slices are stored contiguously in the
reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$ as $n$ seeks are
required to read the slices and $n$ seeks are required to write them:
\begin{equation}
  N_\mathrm{slices} = 2n 
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block. That is, the total number of seeks
performed by Algorithm~\ref{algo:naive-blocks} is:
\begin{equation*}
N_\mathrm{blocks} = n+nd^2 
\end{equation*}
or, using $R$ and $n$ as our main variables:
\begin{equation}
N_\mathrm{blocks} = n+n\left(\sqrt[3]{\frac{R}{n}}\right)^2
\end{equation}
In
practice, this difference could lead to a tremendous slowdown as we will show
later.

@@

Algorithm~\ref{algo:naive-slices} is a particular case of memory
buffering where the amount of available memory equals the maximum size
of a chunk. More buffering can be achieved when the amount of
available memory increases, as shown in
Algorithm~\ref{algo:buffered-slices}.
\begin{algorithm}[h]
  \caption{Buffered merging from slices}
  \label{algo:buffered-slices}
  \begin{algorithmic}[1]
    \STATE sorted\_slices = sort slices by increasing z values
    \STATE initialize buffer
    \FOR{i = 0 ; i \textless n ; i+=1 }
      \STATE slice = sorted\_slices[i]
      \IF{\texttt{sizeof}(buffer)+\texttt{sizeof}(slice) $\geq$ m}
        \STATE write buffer in reconstructed image
        \STATE clear buffer
      \ENDIF
      \STATE read slice and append it to buffer
    \ENDFOR
  \end{algorithmic}
\end{algorithm}
The number of seeks obtained with buffered slices is:
\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}}
\end{equation}
The next two algorithms attempt to generalize buffered slices to blocks.

@@

\subsection{Data}
To benchmark our library we use the 76~GB Big~Brain image split in 125
non-overlapping chunks. The data is available \todo{[here]}.
